{memer}: an R interface to the MEME Suite

### 2020-06-10 manuscript.Rmd
adding more sections, begin rework to use ChIP-seq example to frame discussion.

## Core Utilities
We built a common interface to each MEME Suite tool. Tools requiring sequences
as input can use either a path to a .fasta format file, a
`Biostrings::XStringSet`, or `Biostrings::BStringSetList` as input. These data
structures ensure support for all biological sequence types: DNA, RNA, and
amino-acid. For the remainder of the text, we will focus our examples on DNA
sequences, but `memer` tools support all sequence types. For DNA and RNA
sequences, the `get_sequence()` utility acts as the entrypoint to analysis with
`memer`. `get_sequence()` returns `Biostrings::XStringSet` or
`Biostrings::BStringSetList` types from range-based data types such as
`GenomicRanges` or `GenomicRangesList` which serve as sequence input to the
`memer` core functions.

Some `memer` tools require motifs as input. In these instances, `memer` will
always accept a path to a .meme format file, or a list of motifs in
`universalmotif` format \cite{universalmotif}. The `universalmotif` package
provides simple import and motif conversion functions for all common motif data
structures and file types. Using `universalmotif` objects ensures
cross-compatability with all commonly used motif data formats, providing the
user with maximum flexibility in their analysis. When `memer` functions return
motifs, they are returned to the user in `universalmotif` format. This ensures
that all `memer` motif outputs can be used as inputs to other `memer` functions
that accept motif inputs. Furthermore, `universalmotif` conversion tools allow
`memer` detected motifs to be used in many other popular Bioconductor motif
packages, such as `TFBSTools` and `motifStack`.

The core functionality of `memer` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `runToolname()` function,
which accepts inputs as described above and any valid commandline flags as
function arguments.

```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

```{r, eval=F}
# memer command setting random seed to 100
runDreme(input_sequences, control = "shuffle", s = 100)
```

### Discriminative analysis
```{r}
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")

@@

1. MEME & DREME
 - get_sequences allows parsing genomic coordinates of match from MEME
1. Tools like dreme/meme/ame that can do discriminative testing
2. Tools like tomtom/fimo

### AME/MEME?
 - `get_sequence` solves the problem of having different requirements for fasta format to use certain features
  - score
  - parse_genomic_coord

 - and in some cases extends MEME Suite function to add additional features, such
 as parsing the genomic coordinates of meme matches.
 - also data visualization

@@

## Core Utilities
  - All commandline arguments can be passed to each function, if users mistype the function error handling will suggest typo fixes.
  - `runDreme()`, `runMeme()` = Denovo motif discovery
     - Can be slow
  - `runAme()` = motif enrichment testing
  - `runTomTom()` = motif comparison (often used for matching denovo motifs to known database)
     - both ame & tomtom can use the `meme_db` setting which takes .meme path input, or universalmotif list.
     - can set & forget default in .Renviron
  - `runFimo()` if used with built-in helpers returns genomic coordinates of each motif in input sequences.
     - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
  - for dreme/ame/meme, we've paid special attention to enabling discriminative analysis
     - users can easily perform multiple comparisons of condtion1 vs 2, etc.
     - No special programming is required, no new patterns of data manipulation, etc. Just set input = list_of_sequences, control = "name(s) of entry in list" to compare all entries in list against that set of data (or pooled set if multiple entries).

>>>>

## Core Utilities

The core functionality of `memer` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters.

Tools requiring sequences as input can use a path to a .fasta format
file, a `Biostrings::XStringSet`, or a `Biostrings::BStringSetList` as input.
These data structures support all biological sequence types: DNA,
RNA, and amino-acid. For the remainder of the text, we will 
**use chip-seq data analysis as a framework to highlight memer features**
\strikethrough{focus our examples on DNA sequences.}

For DNA and RNA sequences, the `get_sequence()` utility acts
as the entrypoint to analysis with `memer`. `get_sequence()` returns
`Biostrings::XStringSet` or `Biostrings::BStringSetList` types from range-based
data types such as `GenomicRanges` or `GenomicRangesList` to serve as
sequence input to the `memer` core functions.

When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
Peaks) are commonly represented as `GenomicRanges` objects, which contain
genomic positions of each peak, along with an arbitrary number of metadata
columns describing features of each peak.

Some `memer` tools require motifs as input. In these instances, `memer` will
always accept a path to a .meme format file, or a list of motifs in
`universalmotif` format \cite{universalmotif}. The `universalmotif` package
provides simple import and motif conversion functions for all common motif data
structures and file types, ensuring cross-compatability with all commonly used
motif data formats. When `memer` functions return motifs, they are returned to
the user in `universalmotif` format. This ensures that all `memer` motif outputs
can be used as inputs to other `memer` functions that accept motif inputs.
Furthermore, `universalmotif` conversion tools allow `memer` detected motifs to
be used in many other popular Bioconductor motif packages, such as `TFBSTools`
and `motifStack`, providing the user with maximum flexibility in their analysis.

**Example of commandline vs `memer`?**
For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

The `memer` implementation of the above command would be written as follows:
```{r, eval=F}
# memer command setting random seed to 100
# user explicitly states they want to shuffle input sequences as control
runDreme("input.fasta", control = "shuffle", s = 100)
```

**Somewhere say "memer can be used as a simple wrapper, but here we highlight its utility in chip-seq data analysis".**

### Analysis using multiple groups (split-apply-combine approach)
Users can adopt a "split-apply-combine" strategy for using `memer`. `GRanges`
object acts as the centerpoint, can be split by data, fed to `get_sequence()`,
then the list will run the tools for each group automatically.

### Discriminative analysis
A key feature of several MEME Suite Tools is the ability to perform
discriminative analysis, where motifs are identified or tested for enrichement
in one group of sequences relative to another. In many instances, running tools
in discriminative mode on a set of control sequences outperforms simply using
shuffled input sequences as the control (back this up, change wording?). Therefore, 
enabling complex discriminative analysis designs was central to the design of `memer`.

When using list input to `memer` tools, names of list object members can be
passed to the `control` argument. This will result in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences will be
combined together to create a background set.

```{r}
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")

@@

### Motif Comparison (TomTom)
 - takes runDreme/meme output as input -> pipes

### Motif Matching (FIMO)
 - returns locations as GRanges
 - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa

@@
